\documentclass[preprint]{sigplanconf}

\usepackage{amsmath,amssymb,amsopn,amsthm}
\usepackage[T1]{fontenc}
\usepackage{algorithmic,algorithm}
\usepackage{multirow}
\usepackage{cleveref}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{verbatim}
%\usepackage{stfloats}

\usepackage{tikz}
\usetikzlibrary{fit,positioning}
%\usepackage{pgfplots}
%\usepackage{pgfplotstable}
%\pgfplotsset{compat=1.12} 

\newcommand\todo[1]{{\color{NavyBlue} \footnote{\color{NavyBlue} TODO: {#1}}}}
\newcommand\citneeded[1]{{\color{NavyBlue} \cite{?}\footnote{\color{NavyBlue} citation needed. #1}}}

\newcommand\newterm[1]{{\it #1}}
\newcommand\R{\mathbb{R}}
\newcommand\N{\mathbb{N}}
\newcommand\B{\mathbb{B}}
\newcommand\T{\mathcal{T}}
\renewcommand\S{\mathcal{S}}

\newcommand\lspan[2]{\multicolumn{#1}{@{}l}{#2}}
\newcommand\cspan[2]{\multicolumn{#1}{@{}c}{#2}}

\makeatletter
\newcommand{\LeftEqNo}{\let\veqno\@@leqno}
\makeatother

\newcommand\limplies{\rightarrow}
\newcommand\liff{\leftrightarrow}

\newcommand\exampleTitle{
\bigskip\noindent
\begin{tikzpicture}
  \draw (0,0) -- (5mm,0)
    node(e)[draw,rectangle,rounded corners,right] {\bf Example};
  \draw (e) -- (\columnwidth,0);
\end{tikzpicture}%
\medskip%
}

% author comment macros
\newcommand\authornote[3]{{\color{#2} \footnote{\color{#2} {#1}: {#3}}}}

\newcommand\asolar[1]{\authornote{AS}{BrickRed}{#1}}  % Armando
\newcommand\cel[1]{\authornote{CL}{OliveGreen}{#1}}   % Charles
\newcommand\rch[1]{\authornote{RC}{RoyalPurple}{#1}}  % Rezaul
\newcommand\pga[1]{\authornote{PG}{RawSienna}{#1}}    % Pramod
\newcommand\coa[1]{\authornote{SI}{NavyBlue}{#1}}     % Shachar

\include{macros}

\begin{document}


\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

%\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
%\copyrightyear{20yy} 
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

%\title{Bellmania: Deriving Implementations from Specifications by Transformation}
%\title{Solver-Aided Transformational Synthesis of Divide-and-Conquer Dynamic Programming Algorithms}
\title{Deriving Divide-and-Conquer Dynamic Programming Algorithms Using 
  Solver-Aided Transformations}

%\authorinfo{Shachar Itzhaky}
%           {MIT CSAIL}
%           {shachari@mit.edu}
\authorinfo{Double-blind submission}
           {}
           {}

\maketitle

\begin{abstract}
We introduce \newterm{solver-aided tactics} as a mechanism for transforming
computational terms, expressed in a specialized form of $\lambda$-calculus,
intro equivalent terms in the interest of deriving better, more efficient
implementations. By stringing chains of transformations, a programmer can start
with a compact specification of an algorithm and develop it towards concrete
executable code, while addressing performance issues involved.

Solver-aided tactics are unique in two ways: 
(a) They combine symbolic pattern-matching
with SMT technology to ensure a correct transformation; this means that a lot of the
burden of proof is moved from the user of the system to automated solvers.
(b) They make use of type annotations within program terms to maintain and utilize
properties of those computations, in order to guide transformations and provide
enough context for the automated proofs. For this purpose, we employ predicate
abstraction as part of the type system.

We present Bellmania, a framework that brings together novel techniques from both
inductive and deductive synthesis disciplines.
Bellmania comprises of a language for specifying dynamic programming algorithms as recurrences,
and a calculus that facilitates gradual transformation of these 
specifications into efficient implementations.
In particular, it provides tactics that formalize the Divide-and-Conquer technique to derive
parallel implementations that optimally utilize memory caches.
\end{abstract}

\category{D.1.2}{Programming Techniques}{Automatic Programming}
\category{I.2.2}{Artificial Intelligence}{Automatic Programming}[Program synthesis]

\keywords
Synthesis; Dynamic Programming; SMT; Automated Reasoning; Transformational

\input{01-intro}
\input{02-lang}
\input{03-tactics}
\input{04-automated}
\input{05-evaluation}
\input{06-related}
\input{07-conc}

\appendix

\input{A-more-tactics}

\bibliographystyle{abbrvnat}
\bibliography{popl2016} % embed contents of popl2016.bbl for source submission


\end{document}
